---
title: "R Packages Development"
author: "Hamoud Alhoqbani"
date: "1/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Short manual for creating your own package.

## Step 1: Prepare development environment

### Install required/recommended packages

- devtools (automate common development tasks)
- roxygen2 (generate documentation)
- testthat (automated tests)
- knitr (report generation for Vignettes)

```{r install packages, message=FALSE, warning=FALSE}
# install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
if(!require(devtools)) install.packages("devtools",repos = "http://cran.us.r-project.org")
if(!require(roxygen2)) install.packages("roxygen2",repos = "http://cran.us.r-project.org")
if(!require(testthat)) install.packages("testthat",repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("knitr",repos = "http://cran.us.r-project.org")
```


### Verify environment is ready for packages development
```{r check environment}
library(devtools)
has_devel() # This should return TRUE
```


## Step 2: Create new package using RStudio
You can use `devtools::create()` to create project skeleton.
However, it's easier to use RStudio to create the project.

### Create the package files using RStudio

1. File > New Project > New Directory >  R Package
2. Fill the details: 
    - choose a name: package name requirements:
        * can have only: letters, numbers and periods
        * must start with a letter
        * cannot end with a period
    - location where the project file should be saved.
    - Check enable  git repository.
    - Check open in new session.


### Add and commit the files to git repository
```
git status
git add .
git status
git commit -m 'Initial commit'
git status
```

## Step 3: Write Package Functions

All custom R scripts for the package should be stored in the `R` directory.

- Create a new R script and save it to the R folder inside the package project directory. 
- You can choose any name to the file, but it's recommended to choose a name that describes the purpose of the functions inside the file. 
- You can put each function in a separate file, or groups related functions into one file.

## Step 4: Using Your Package
Before using the package, you must install it into your `library`.

A package can be installed using different methods. 

The most common method is installing the package from cran repository using `install.packages()`

Using the `devtools` package, you can install packages from github.com or from a local directory.
or from other sources.


### Installing the package

Because we are not going to upload the package to `cran`, we will use github. 
However, we will test installing the package locally first.


#### Options 1: Loading the package from local location

From you project, e.g. (r-Practice), load your new package to use it.
```{r}
devtools::install_local('path/to/package')
```

#### Option 2: Loading the package from github.com

1. Add your new files and all changes to git:
```
git status
git add .
git commit -m 'Add pacakge custom functions'
```
2. Create a repository in github and push the package:
Visit https://github.com and create new repository and give it the same name as your package name. 

The address of the new repo will be something like: https://github.com/username/packagename

Upload/Push your package files to the new repository.
```
git remote add origin git@github.com:username/packagename.git
git push -u origin master
```

3. Install the package to your library.
```{r}
devtools::install_github('username/packagename')
```

### Loading and using the package
```
library(packagename)
package.function()
# or using the namespace
packagename::package.function()
```


# Resources

- http://r-pkgs.had.co.nz/
- https://www.datacamp.com/community/tutorials/r-packages-guide